Linear instability underlying synchronous waving of marine grass in flow

### 2014-09-21 Paper/grass.tex
minor update

\begin{abstract}
We show that the threshold flow condition for the spontaneous waving of marine grass arises due to the drag exerted by the vegetation on the flow. 
We show this by developing a self-consistent model of the flow past rigid vegetation. 
Linear stability analysis of the steady flow resulting from this model naturally leads to the threshold condition, and explains its mechanistic origin. 
Asymptotic analysis of the model for dense grass reveals two unstable modes. 
One of these modes shares many common features with the Kelvin-Helmholtz instability, but is demonstrated to be different from it. 
The predictions of the model for threshold condition and the resulting dominant frequency of waving agree with laboratory observations. 
\end{abstract}
\maketitle
%\section{Introduction}
Sea grasses exhibit a rich set of dynamical behavior due to their
 collective interaction with both steady and oscillatory flows. 
%interaction with the flow of water and can considerably affect surrounding hydrodynamic conditions.
%Resulting 
The resulting changes in hydrodynamic conditions can influence number of environmental processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}.
The simplest response of these flexible grass canopies 
%Flexible grass canopies respond 
to steady currents 
%in the form of 
is the formation of coherent large amplitude oscillations, known as \monami \cite{AckermanOkubo93}. 
%Such motion is considered to play a central role in the recruitment 
%of microscopic marine organisms, such as blue mussel larvae\cite{Grizzle96}.
%The hydrodynamic mechanism underlying \monami is the focus of this paper. 
In this letter we provide a hydrodynamic mechanism for the onset of these coherent oscillations.

%Existing 
Current explanations of \monami invoke the existence of a shear layer near the top of the canopy \cite{Ghisal02,Raupach96} and its instability 
%due to different amounts of drag experienced by fluid in and above the canopy. 
%This shear layer is interpreted to become unstable to coherent vortices 
through a mechanism similar to the Kelvin-Helmholtz instability. This is thought to lead to coherent eddies over the sea grass canopy which then drives 
%Influence of these coherent eddies over sea grasses is manifested in their 
large amplitude synchronous oscillations.

@@

While the shear layer model appears to predict the frequency of \monami consistent with experimental observations, several aspect of existing theory remain unsatisfactory. 
(i) While the free shear layer is established due to the vegetation drag, linear stability analysis\cite{Raupach96} does not account for this, rendering this explanation inconsistent. 
(ii) Classical free shear flow is known to be unstable for all Reynolds number\cite{drazin}, whereas observations in the lab\cite{Ghisal02} and the field\cite{Grizzle96} indicate the existence of a threshold 
flow speed below which \monami is not observed. 
(iii)The thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the grass tip, is in many cases comparable to the unvegetated layer thickness,
and therefore inconsistent with the assumptions of the free shear layer instability. 
% These drawbacks of existing theory suggest that flow through vegetation requires further investigation for better understanding of the phenomenon.

Here we present a self-consistent mathematical model that account correctly for these effects while also explaining 
%exhibiting a rich behavior, which is not only consistent within itself, but also with 
lab experiments and field observations.
%Using this model, the synchronous waving may be understood to be arising from the linear hydrodynamic instability of a uniform flow over the grass bed, such that the principal role of the vegetation is to exert a drag on the flow. 
%A threshold flow speed arises naturally for the onset of the instability; this threshold condition and the resulting waving frequency compares well with experimental observations.
%The vegetation drag introduces a damping mechanism in the flow, in addition to the viscous damping, and overcoming this damping is the primary consideration that sets the threshold flow.
%We deduce this numerically and through an asymptotic scaling analysis in the limit of dense grass.
\newline
The drag exerted by the grass bed on the flow is central to the hypothesized mechanism\cite{Ghisal02} of the instability underlying \monami because it realizes the shear layer structure of the flow. 
The flexibility of the grass blades, on the other hand, serves merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 
It has been demonstrated that the instability and the resulting flow structures are present even when the grass blades are replaced with rigid dowels\cite{Ghisal02}. 
Therefore, for simplicity, 
%mathematical model for the linear instability, 
we consider the grass bed to form a rigid structure that exchanges momentum with the surrounding fluid but does not deform. 
Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated\cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, 
the flow instability is similar to the instability of a shear layer.

@@

To understand the onset of, we first calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies 
\begin{equation}
 -{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0
\label{base_equ}
\end{equation}
where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$. In the limit of large grass density, shear stress exerted by the bottom surface is estimated to be negligible in comparison with the drag and hence the bottom surface ($y=0$) is approximated by zero shear.   
%The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, 
%as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
A comparison of laboratory mean profile with the steady flow profile is shown in figure \eqref{basicflow}.
Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and with pressure gradient. Flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 
Continuity of flow speed and shear stresses at canopy top gives rise to a boundary layer 
%in the flow profile 
near the tip of the grass.

>>>>

\begin{abstract}
We show that the threshold flow condition for the spontaneous waving of marine grass arises due to the drag exerted by the vegetation on the flow. 
We show this by developing a self-consistent model of the flow past rigid vegetation. 
Linear stability analysis of the steady flow resulting from this model naturally leads to the threshold condition, and explains its mechanistic origin. 
Asymptotic analysis of the model for dense grass reveals two unstable modes. 
One of these modes shares many common features with the Kelvin-Helmholtz instability, but is demonstrated to be different from it. 
The predictions of the model for threshold condition and the resulting dominant frequency of waving agree with laboratory observations. 
\end{abstract}
\maketitle
%\section{Introduction}
Sea grasses exhibit a rich set of dynamical behavior due to their collective interaction with both steady and oscillatory flows.  
The resulting changes in hydrodynamic conditions can influence number of environmental processes such as transport of sediments, contaminants, dissolved oxygen, plant growth, and biomass production \cite{Fonseca87,Grizzle96,Nepf99,Nepf2012}. The simplest response of these flexible grass canopies to steady currents is the formation of coherent large amplitude oscillations, known as \monami \cite{AckermanOkubo93}. 
%Such motion is considered to play a central role in the recruitment 
%of microscopic marine organisms, such as blue mussel larvae\cite{Grizzle96}.
%The hydrodynamic mechanism underlying \monami is the focus of this paper. 
In this letter we provide a hydrodynamic mechanism for the onset of these coherent oscillations.
Current explanations of \monami invoke the existence of a shear layer near the top of the canopy \cite{Ghisal02,Raupach96} and its instability 
%due to different amounts of drag experienced by fluid in and above the canopy. 
%This shear layer is interpreted to become unstable to coherent vortices 
through a mechanism similar to the Kelvin-Helmholtz instability. This is thought to lead to coherent eddies over the sea grass canopy which then drives 
%Influence of these coherent eddies over sea grasses is manifested in their 
large amplitude synchronous oscillations.

@@

While the shear layer model appears to predict the frequency of \monami consistent with experimental observations, several aspect of existing theory remain unsatisfactory. 
(i) While the free shear layer is established due to the vegetation drag, linear stability analysis\cite{Raupach96} does not account for this, rendering this explanation inconsistent. 
(ii) Classical free shear flow is known to be unstable for all Reynolds number\cite{drazin}, whereas observations in the lab\cite{Ghisal02} and the field\cite{Grizzle96} indicate the existence of a threshold 
flow speed below which \monami is not observed. 
(iii)The thickness of the free shear layer, defined as the momentum thickness of the boundary layer near the grass tip, is in many cases comparable to the unvegetated layer thickness,
and therefore inconsistent with the assumptions of the free shear layer instability.
Here we present a self-consistent mathematical model that account correctly for these effects 
while also explaining lab experiments and field observations.
%Using this model, the synchronous waving may be understood to be arising from the linear hydrodynamic instability of a uniform flow over the grass bed, such that the principal role of the vegetation is to exert a drag on the flow. 
%A threshold flow speed arises naturally for the onset of the instability; this threshold condition and the resulting waving frequency compares well with experimental observations.
%The vegetation drag introduces a damping mechanism in the flow, in addition to the viscous damping, and overcoming this damping is the primary consideration that sets the threshold flow.
%We deduce this numerically and through an asymptotic scaling analysis in the limit of dense grass.
\newline
The drag exerted by the grass bed on the flow is central to the hypothesized mechanism\cite{Ghisal02} of the instability underlying \monami because it realizes the shear layer structure of the flow. 
The flexibility of the grass blades, on the other hand, serves merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 
It has been demonstrated that the instability and the resulting flow structures are present even when the grass blades are replaced with rigid dowels\cite{Ghisal02}. 
Therefore, for simplicity, 
%mathematical model for the linear instability, 
we consider the grass bed to form a rigid structure that exchanges momentum with the surrounding fluid but does not deform. 
Indeed, a more detailed calculation undertaken for terrestrial grass demonstrated\cite{Delangre06} that when the timescale of the instability is well-separated from the period of natural oscillations of the grass bed, 
the flow instability is similar to the instability of a shear layer.

@@

To understand the onset of \monami, we first calculate the steady flow $\bu = U(y)\boldsymbol{\hat{x}}$ as a solution of \eqref{navier-stokes}, and examine its linear stability. 
Such a steady state flow driven by constant pressure gradient satisfies 
\begin{equation}
 -{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0
\label{base_equ}
\end{equation}
where $S(y)=1$ in the vegetated layer $0<y<\hg$ and $S(y)=0$ in the unvegetated layer $\hg< y< 2H$. In the limit of large grass density, shear stress exerted by the bottom surface is estimated to be negligible in comparison with the drag and hence the bottom surface ($y=0$) is approximated by zero shear.   
%The flow in the laboratory experiments is better approximated by zero shear at the bottom ($y=0$) and top ($y=2H$) of the fluid layer, 
%as can be seen from the comparison between sample solutions and mean flow observed in laboratory experiment shown in Figure \ref{basicflow}. 
A comparison of laboratory mean profile with the steady flow profile is shown in figure \eqref{basicflow}.
Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and with pressure gradient. Flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 
Continuity of flow speed and shear stresses at canopy top gives rise to a boundary layer 
%in the flow profile 
near the tip of the grass.
